Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-08 method.tex
spell check
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@980 ccdd05d5-56c1-461a-bf98-85f6876523b4

In this section we describe a decomposition of the \CBC{} waveform space that
reduces time-domain (\TD) filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in
a search with minutes latency in \LIGO{}'s S6 and Virgo's VSR3 joint science
runs~\cite{HugheyGWPAW2011}.  We combine this with the \SVD\ rank-reduction
method of~\cite{Cannon:2010p10398} that exploits the redundancy of
the template banks.

@@

Here we explore a method for reducing the computational cost of a \TD\ search
for compact binary coalescence.  We will give a truly zero-latency algorithm
that competes in terms of floating point operations per second with the
conventional \FD\ method, which requires a signficant latency in order to be
computationally cheap. Our method, called \lloid{} (a backronym for
Low Latency Online Inspiral Detection),
involves two transformations of the template waveforms that produce a set of
orthogonal filters with far fewer coefficients than the original templates.

The first transformation is to chop the \TD{} templates up into \emph{time
slices}.  Since each template slice is disjoint in time, the resulting set for
a single template is orthogonal.  Given the chirp-like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely downsampled.  The downsampling reduces the number of data samples by a
factor of $\sim$100 from the early part of the waveform and allows the filters
to be evaluated at about $\sim$$1/100^\mathrm{th}$ the sample rate. This
amounts to more than a factor of 10000
reduction in the floating-point operations per second required to filter the
early part of the waveform.  However, the resulting filters are still not
orthogonal across the mass parameter space, and are in fact highly redundant.
We use the \SVD{} to produce an orthogonal filter set
from the time-sliced templates~\cite{Cannon:2010p10398}.  We find that this
reduces the number of sample points of the filters by another factor of
$\sim$100.  The combined methods reduce the number of floating point operations
to the level where they are competitive with the conventional high-latency \FD\
matched filter approach.  In the remainder of this section we describe the
\lloid\ algorithm in detail and provide some basic computational cost scaling.

@@

For concreteness and simplicity, we will consider an inspiral waveform in the
quadrupole approximation, for which the time-frequency relation is
%
\begin{equation} \label{eq:fgw}
%
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \umu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary~\cite{findchirppaper, kidder1992}.  Usually the
template is truncated at some prescribed time $t^0$, or equivalently frequency
$f_\mathrm{hi}$.  This is often chosen to correspond to the \ISCO. An inspiral
signal will enter the detection band at a low frequency, $f = f_\mathrm{low}$,
corresponding to a time $t_\mathrm{low}$.  The template is assumed to be zero
outside the interval $[t^0, t_\mathrm{low})$ and is said to have  a duration of
$t_\mathrm{low} - t^0$. It is critically sampled at a rate of $2
f_\mathrm{hi}$.

The monotonic time-frequency relationship of equation~\eqref{eq:fgw} allows us
to choose time-slice boundaries that require substantially less bandwidth at
early times in the inspiral.  Our goal is to reduce the filtering cost of a
large fraction of the waveform by computing part of the convolution at a lower
sample rate.  Specifically we consider here time slice boundaries with the
smallest power-of-two sample rate that sub-critically samples the time-sliced
template.  The time slices consist of the $S$ intervals
$\left[t^0, t^1\right),\, \left[t^1, t^2\right),\, \dots,\, \left[t^{S-1}, t^S\right)$,
sampled at frequencies $f^0,\, f^1,\, \dots,\, f^\mathrm{S-1}$ where $f^0
\geqslant 2 f_\mathrm{hi}$ and $f^{S-1} \geqslant 2 f_\mathrm{low}$.  The time
sliced templates may be downsampled without aliasing, so we define them as
%
\begin{equation}
\label{eq:time-sliced-templates}
h_{i}^{s}[k] \equiv
	\cases{
		h_{i}\!\left[k\frac{f}{f^s}\right] & if $t^s \leqslant k/f^s < t^{s+1}$ \\
		0 & otherwise.
	}
\end{equation}
%
\begin{comment}
An example time-slice design satisfying these constraints for a $1.4 - 1.4 \,
M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\vspace{0.8cm}
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{hi} = f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time-frequency structure given by ($\ref{eq:fgw})$. $f^s$ is the sample
rate of the time slice, $(t^{s+1}, t^s]$ are the boundaries in seconds
preceeding coalescence and \slicessamps\ is the number of samples in the
$s^{\mathrm{th}}$ filter.}
\end{table}
\end{comment}

@@

In the previous two sections we have described two transformations that greatly
reduce the burden of filtering waveforms from the the compact binary parameter space and make \TD\
filtering of the data possible.  We are now prepared to define the early
warning \SNR\ and to comment on the computational cost of evaluating it.  But
first, we will introduce some notation refering to the decimation of data,
which means reducing the sample rate after low-pass filtering,
\begin{equation}
\label{eq:decomp}
	x^{s+1}[k] = \left( H^\shortdownarrow x^s\right)[k] \,,
\end{equation}
and notation refering to the interpolation of data to increase the sample rate,
\begin{equation}
	x^{s}[k] = \left( H^\shortuparrow x^{s+1}\right)[k] \,.
\end{equation}

From the combination of transformations to the compact binary templates defined
in equation \eqref{eq:time-sliced-templates} and \eqref{eq:svddecomp} we define
the early-warning filter output accumulated up to time slice $s$ as,
%
% orthogonal decomposition filtering
%
\begin{equation}
	\rho_i^s [k] =%
		% Reconstruction
		\underbrace{
			\sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s
		}_\textrm{\clap{reconstruction}}
		% Orthogonal FIR filter
		\overbrace{
			\sum_{\mathclap{n=0}}^{\mathclap{N^s-1}} u_l^s[n] x^s[k-n]
		}^\textrm{\clap{orthogonal {\sc fir} filters}}
		% Plus ...
		+
		% Interpolation SNR
		\underbrace{
			\left(H^\uparrow \rho_i^{s+1}\right)[k]
		}_\textrm{\clap{{\sc snr} from previous time slices}}
\end{equation}
%
%
This quantity is related to the early-warning \SNR{} by the normalization
of the partial filter defined by the present time slice, which may be computed
in advance. \editorial{Chad: are we happy with that statement?}
%
%
\begin{figure}[htbp]
	\includegraphics{figures/lloid-diagram.pdf}
	\caption{\label{fig:pipeline} Schematic of \lloid{} pipeline illustrating
signal flow.  Circles with arrows represent interpolation
\protect\includegraphics{figures/upsample-symbol.pdf} or decimation
\protect\includegraphics{figures/downsample-symbol.pdf}.  Circles with plus
signs represent summing junctions
\protect\includegraphics{figures/adder-symbol.pdf}.  Squares
\protect\includegraphics{figures/fir-symbol.pdf} stand for \fir{} filters.  Sample
rate decreases from the top of the diagram to the bottom.  In this diagram each
time slice contains three \fir\ filters that are linearly combined to produce
four output channels.  In a typical pipeline the number of \fir\ filters is
much less than the number of output channels.}
\end{figure}
%
%
The signal flow diagram in figure~\ref{fig:pipeline} illustrates how this
recursion relation can be realized as a multirate filter network with outputs for
each of the early-warning \SNR{}s.

@@

The computational cost of the decimation of the detector data is a little bit more subtle.  Decimation is achieved by applying an \fir\ anti-aliasing filter and then downsampling, or deleting samples in order to reduce the sample rate from $f^{s-1}$ to $f^s$.  Naively, an anti-aliasing filter with $N^\shortdownarrow$ coefficients should demand $2 N^\shortdownarrow f^{s-1}$ \flops.  However, it is necessary to evaluate the anti-aliasing filter only for the fraction $f^s / f^{s-1}$ of the samples that will not be deleted.  Consequently, an efficient decimator that requires only $2 N^\shortdownarrow f^{s-1} \cdot \left( f^s / f^{s-1} \right) = 2 N^\shortdownarrow f^s$ \flops\ is possible.

>>>>

In this section we describe a decomposition of the \CBC{} waveform space that
reduces time-domain (\TD) filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multi-band decomposition of the compact binary parameter space that resulted in
a search with minutes latency in \LIGO{}'s S6 and Virgo's VSR3 joint science
runs~\cite{HugheyGWPAW2011}.  We combine this with the \SVD\ rank-reduction
method of~\cite{Cannon:2010p10398} that exploits the redundancy of
the template banks.

@@

Here we explore a method for reducing the computational cost of a \TD\ search
for compact binary coalescence.  We will give a truly zero-latency algorithm
that competes in terms of floating point operations per second with the
conventional \FD\ method, which requires a significant latency in order to be
computationally cheap. Our method, called \lloid{} (a backronym for
Low Latency Online Inspiral Detection),
involves two transformations of the template waveforms that produce a set of
orthogonal filters with far fewer coefficients than the original templates.

The first transformation is to chop the \TD{} templates up into \emph{time
slices}.  Since each template slice is disjoint in time, the resulting set for
a single template is orthogonal.  Given the chirp-like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely down-sampled.  The down-sampling reduces the number of data samples by a
factor of $\sim$100 from the early part of the waveform and allows the filters
to be evaluated at about $\sim$$1/100^\mathrm{th}$ the sample rate. This
amounts to more than a factor of 10000
reduction in the floating-point operations per second required to filter the
early part of the waveform.  However, the resulting filters are still not
orthogonal across the mass parameter space, and are in fact highly redundant.
We use the \SVD{} to produce an orthogonal filter set
from the time-sliced templates~\cite{Cannon:2010p10398}.  We find that this
reduces the number of sample points of the filters by another factor of
$\sim$100.  The combined methods reduce the number of floating point operations
to the level where they are competitive with the conventional high-latency \FD\
matched filter approach.  In the remainder of this section we describe the
\lloid\ algorithm in detail and provide some basic computational cost scaling.

@@

For concreteness and simplicity, we will consider an inspiral waveform in the
quadrapole approximation, for which the time-frequency relation is
%
\begin{equation} \label{eq:fgw}
%
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \umu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary~\cite{findchirppaper, kidder1992}.  Usually the
template is truncated at some prescribed time $t^0$, or equivalently frequency
$f_\mathrm{hi}$.  This is often chosen to correspond to the \ISCO. An inspiral
signal will enter the detection band at a low frequency, $f = f_\mathrm{low}$,
corresponding to a time $t_\mathrm{low}$.  The template is assumed to be zero
outside the interval $[t^0, t_\mathrm{low})$ and is said to have  a duration of
$t_\mathrm{low} - t^0$. It is critically sampled at a rate of $2
f_\mathrm{hi}$.

The monotonic time-frequency relationship of equation~\eqref{eq:fgw} allows us
to choose time-slice boundaries that require substantially less bandwidth at
early times in the inspiral.  Our goal is to reduce the filtering cost of a
large fraction of the waveform by computing part of the convolution at a lower
sample rate.  Specifically we consider here time slice boundaries with the
smallest power-of-two sample rate that sub-critically samples the time-sliced
template.  The time slices consist of the $S$ intervals
$\left[t^0, t^1\right),\, \left[t^1, t^2\right),\, \dots,\, \left[t^{S-1}, t^S\right)$,
sampled at frequencies $f^0,\, f^1,\, \dots,\, f^\mathrm{S-1}$ where $f^0
\geqslant 2 f_\mathrm{hi}$ and $f^{S-1} \geqslant 2 f_\mathrm{low}$.  The time
sliced templates may be down-sampled without aliasing, so we define them as
%
\begin{equation}
\label{eq:time-sliced-templates}
h_{i}^{s}[k] \equiv
	\cases{
		h_{i}\!\left[k\frac{f}{f^s}\right] & if $t^s \leqslant k/f^s < t^{s+1}$ \\
		0 & otherwise.
	}
\end{equation}
%
\begin{comment}
An example time-slice design satisfying these constraints for a $1.4 - 1.4 \,
M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\vspace{0.8cm}
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{hi} = f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time-frequency structure given by ($\ref{eq:fgw})$. $f^s$ is the sample
rate of the time slice, $(t^{s+1}, t^s]$ are the boundaries in seconds
preceding coalescence and \slicessamps\ is the number of samples in the
$s^{\mathrm{th}}$ filter.}
\end{table}
\end{comment}

@@

In the previous two sections we have described two transformations that greatly
reduce the burden of filtering waveforms from the the compact binary parameter space and make \TD\
filtering of the data possible.  We are now prepared to define the early
warning \SNR\ and to comment on the computational cost of evaluating it.  But
first, we will introduce some notation referring to the decimation of data,
which means reducing the sample rate after low-pass filtering,
\begin{equation}
\label{eq:decomp}
	x^{s+1}[k] = \left( H^\shortdownarrow x^s\right)[k] \,,
\end{equation}
and notation referring to the interpolation of data to increase the sample rate,
\begin{equation}
	x^{s}[k] = \left( H^\shortuparrow x^{s+1}\right)[k] \,.
\end{equation}

From the combination of transformations to the compact binary templates defined
in equation \eqref{eq:time-sliced-templates} and \eqref{eq:svddecomp} we define
the early-warning filter output accumulated up to time slice $s$ as,
%
% orthogonal decomposition filtering
%
\begin{equation}
	\rho_i^s [k] =%
		% Reconstruction
		\underbrace{
			\sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s
		}_\textrm{\clap{reconstruction}}
		% Orthogonal FIR filter
		\overbrace{
			\sum_{\mathclap{n=0}}^{\mathclap{N^s-1}} u_l^s[n] x^s[k-n]
		}^\textrm{\clap{orthogonal {\sc fir} filters}}
		% Plus ...
		+
		% Interpolation SNR
		\underbrace{
			\left(H^\uparrow \rho_i^{s+1}\right)[k]
		}_\textrm{\clap{{\sc snr} from previous time slices}}
\end{equation}
%
%
This quantity is related to the early-warning \SNR{} by the normalization
of the partial filter defined by the present time slice, which may be computed
in advance. \editorial{Chad: are we happy with that statement?}
%
%
\begin{figure}[htbp]
	\includegraphics{figures/lloid-diagram.pdf}
	\caption{\label{fig:pipeline} Schematic of \lloid{} pipeline illustrating
signal flow.  Circles with arrows represent interpolation
\protect\includegraphics{figures/upsample-symbol.pdf} or decimation
\protect\includegraphics{figures/downsample-symbol.pdf}.  Circles with plus
signs represent summing junctions
\protect\includegraphics{figures/adder-symbol.pdf}.  Squares
\protect\includegraphics{figures/fir-symbol.pdf} stand for \fir{} filters.  Sample
rate decreases from the top of the diagram to the bottom.  In this diagram each
time slice contains three \fir\ filters that are linearly combined to produce
four output channels.  In a typical pipeline the number of \fir\ filters is
much less than the number of output channels.}
\end{figure}
%
%
The signal flow diagram in figure~\ref{fig:pipeline} illustrates how this
recursion relation can be realized as a multi-rate filter network with outputs for
each of the early-warning \SNR{}s.

@@

The computational cost of the decimation of the detector data is a little bit more subtle.  Decimation is achieved by applying an \fir\ anti-aliasing filter and then down-sampling, or deleting samples in order to reduce the sample rate from $f^{s-1}$ to $f^s$.  Naively, an anti-aliasing filter with $N^\shortdownarrow$ coefficients should demand $2 N^\shortdownarrow f^{s-1}$ \flops.  However, it is necessary to evaluate the anti-aliasing filter only for the fraction $f^s / f^{s-1}$ of the samples that will not be deleted.  Consequently, an efficient decimator that requires only $2 N^\shortdownarrow f^{s-1} \cdot \left( f^s / f^{s-1} \right) = 2 N^\shortdownarrow f^s$ \flops\ is possible.
